---
title: "Module-6"
format: html
editor: visual
---

# Module 6

```{r}
library(tidyverse)
```

# Challenge 1

Load the Country-Data-2016 into a data frame varaible, d, and summarize the variables in that data frame. You can load the .csv file with a link or from local file so I'm going to try both.

Load with link/curl

```{r}
library(curl)
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/Country-Data-2016.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
```

## OR

Load from local file

```{r}
f <- "~/Desktop/Country-Data-2016.csv"
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
```

```{r}
summary(d)
```

```{r}
names(d)
```

Create a new pop_density variable in your data frame which is population / area. What are the 10 most dense countries? The 10 least dense?

Most

```{r}
d$pop_density <- d$population/d$area
d <- d[order(-d$pop_density), ]
d[1:10, ]
```

Least

```{r}
d <- d[order(d$pop_density), ]
d[1:10, ]
```

Extract data from the 20 largest countries into a new variable (I believe a new varaible refers to a new dataframe (NOT d)). What are the median area and population size of these countries?

```{r}
big_20 <- d[order(-d$area), ] [1:20,]
summary(big_20)
```

## OR

```{r}
mean(new$population, na.rm = TRUE)
mean(new$area, na.rm = TRUE)
```

Extract data from all countries beginning with the letters “A” through “F”. What are the mean area and population size of these countries?

grep looks for A-F anywhere in the name of the country

```{r}
a_to_f <- d[grep("[A-F]", d$country), ]
summary(a_to_f)
```

# Challenge 2

```{r}
par(mfrow = c(2, 3))
boxplot(d$population)
boxplot(log(d$population))
boxplot(d$area)
boxplot(log(d$area))
barplot(d$population)
barplot(d$area)
```

# Challenge 3

Make histograms of the log() transformed population and area data from the Country-Data-2016 file. Explore what happens if you set freq=FALSE versus the default of freq=TRUE. Try looking at other variables as well.

Setting the argument freq=FALSE will scale the Y axis to represent the proportion of observations falling into each bin rather than the count.

```{r}
par(mfrow = c(1, 2))  # gives us two panels
attach(d)
hist(log(population), freq = FALSE, col = "red", main = "Plot 1", xlab = "log(population size)",
    ylab = "density", ylim = c(0, 0.2))
hist(log(area), freq = FALSE, col = "red", main = "Plot 2", xlab = "log(area)",
    ylab = "density", ylim = c(0, 0.2))
```

NOTE: You can add a line to your histograms (e.g., to show the mean value for a variable) using the abline() command, with arguments. For exmaple, to show a single vertical line representing the mean log(population size), you would add the argument v=mean(log(population)))

Density plot : The density() function computes a non-parametric estimate of the distribution of a variable, which can be combined with plot() to also yield a graphical view of the distribution of the data. If your data have missing values, then you need to add the argument na.rm=TRUE to the density() function. To superimpose a density() curve on a histogram, you can use the lines(density()) function.

```{r}
par(mfrow = c(1, 1))  # set up one panel and redraw the log(population) histogram
hist(log(population), freq = FALSE, col = "white", main = "My Plot with Mean and Density",
    xlab = "log(population size)", ylab = "density", ylim = c(0, 0.2))
abline(v = mean(log(population), na.rm = TRUE), col = "blue")
lines(density(log(population), na.rm = TRUE), col = "green")
```

```{r}
detach(d)
```

Tables: the table() function can be used to summarize counts and proportions for categorical variables in your dataset.

Using the table() function, find what is the most common form of government in the Country-Data-2016 dataset. How many countries have that form? HINT: We can combine table() with sort() and the argument decreasing=TRUE to get the desired answered straight away:

```{r}
sort(table(d$govt_form), decreasing = TRUE)
```

# Challenge 5

Load KamilarAndCooper data

```{r}
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/KamilarAndCooperData.csv")
d <- read.csv(f, header = TRUE, stringsAsFactors = FALSE)
attach(d) # attaching means you won't have to use $ symbol?
head(d)
summary(d)
```

Make boxplot

```{r}
boxplot(log(Body_mass_female_mean) ~ Family, d) # ~Family means to group by family
```

Don't forget to detach!

```{r}
detach(d)
```

## Alternative: use ggplot2 (could be done as 1 line but shown as multiple for readability)

```{r}
library(ggplot2)
p <- ggplot(data = d, aes(x = Family, y = log(Body_mass_female_mean)))  #define the variables
p <- p + geom_boxplot()  #graph them in a boxplot
p <- p + theme(axis.text.x = element_text(angle = 90))  #put x-axis names at 90deg
p <- p + ylab("log(Female Body Mass)")  #rename y-axis title
p  #show me the graph
```

# Challenge 6

Scatterplots : Scatterplots are a natural tool for visualizing two continuous variables and can be made easily with the plot(x=XXX, y=YYY) function in {base} graphics (where XXX\* and YYY\*\* denote the names of the two variables you wish to plot). Transformations of the variables, e.g., log or square-root (sqrt()), may be necessary for effective visualization.

Again using data from the KamilarAndCooperData dataset, plot the relationship between female body size and female brain size. Then, play with log transforming the data and plot again.

```{r}
attach(d)
par(mfrow = c(1, 2))
plot(x = Body_mass_female_mean, y = Brain_Size_Female_Mean)
plot(x = log(Body_mass_female_mean), y = log(Brain_Size_Female_Mean))
```

```{r}
detach(d)
```

## Alternative: ggplot2

```{r}
p <- ggplot(data = d, aes(x = log(Body_mass_female_mean), y = log(Brain_Size_Female_Mean),
    color = factor(Family)))  # first, we build a plot object and color points by Family
p <- p + xlab("log(Female Body Mass)") + ylab("log(Female Brain Size)")  # then we modify the axis labels
p <- p + geom_point()  # then we make a scatterplot
p <- p + theme(legend.position = "bottom", legend.title = element_blank())  # then we modify the legend
p  # and, finally, we plot the object
```

Using {ggplot2}, we can also easily set up a grid for “faceting”" by a grouping variable

```{r}
p <- p + facet_wrap(~Family, ncol = 4)
p <- p + theme(legend.position = "none")
p
```

Add regression to plot

```{r}
p <- p + geom_smooth(method = "lm", fullrange = TRUE)
p
```

# Challenge 7

```{r}
p <- ggplot(data = d, aes(x = log(Body_mass_female_mean), y = log(Mass_Dimorphism)))
p <- p + geom_point()
p <- p + geom_smooth(method = "lm")
p
```

To calculate summary statistics for groups of observations in a data frame, there are many different approaches. One is to use the aggregate() function from the {stats} package (a standard package), which provides a quick way to look at summary statistics for sets of observations, though it requires a bit of clunky code. Here, we apply a particular function (FUN = "mean") to mean female body mass, grouped by Family.

```{r}
aggregate(d$Body_mass_female_mean ~ d$Family, FUN = "mean", na.rm = TRUE)
```

## OR

```{r}
aggregate(x = d["Body_mass_female_mean"], by = d["Family"], FUN = "mean", na.rm = TRUE)
```

### dplyr package how to

```{r}
library(dplyr)
```

```{r}
s <- filter(d, Family == "Hominidae" & Mass_Dimorphism > 2)
head(s)  # filtering a data frame for certain rows...
```

```{r}
s <- arrange(d, Family, Genus, Body_mass_male_mean)  # rearranging a data frame...
head(s)
```

```{r}
s <- select(d, Family, Genus, Body_mass_male_mean)  # selecting specific columns...
head(s)
```

```{r}
s <- rename(d, Female_Mass = Body_mass_female_mean)
head(s$Female_Mass)  # renaming columns...
```

```{r}
s <- mutate(d, Binomial = paste(Genus, Species, sep = " "))
head(s$Binomial)  # and adding new columns...
```

```{r}
s <- summarise(d, avgF = mean(Body_mass_female_mean, na.rm = TRUE), avgM = mean(Body_mass_male_mean,
    na.rm = TRUE))
s
```

```{r}
byFamily <- group_by(d, Family)
byFamily
```

```{r}
s <- summarise(byFamily, avgF = mean(Body_mass_female_mean, na.rm = TRUE), avgM = mean(Body_mass_male_mean,
    na.rm = TRUE))
s
```

## Piping

You can pipe instead of nest operations Below accomplishes what the previous line did

```{r}
s <-                  #to create dataframe "s"
  d %>%                 #take dataframe "d"
  group_by(Family) %>%  #Group it by Family
  summarise(avgF = mean(Body_mass_female_mean, na.rm=TRUE),      #And calculate mean male BM
            avgM = mean(Body_mass_male_mean, na.rm=TRUE))        #And mean female BM
s
```

### Challenge 8

Add a variable, Binomial to our data frame d, which is a concatenation of the Genus and Species… Trim the data frame to only include the variables Binomial, Family, Body_mass_female_mean, Body_mass_male_mean and Mass_Dimorphism… Group these by Binomial (in other words, by species but with full binomial nomenclature)… And calculate the average value for female body mass, male body mass, and mass dimorphism.

```{r}
s <- d %>%
    mutate(Binomial = paste(Genus, Species, sep = " ")) %>%
    select(Binomial, Body_mass_female_mean, Body_mass_male_mean, Mass_Dimorphism) %>%
    group_by(Binomial) %>%
    summarise(avgF = mean(Body_mass_female_mean, na.rm = TRUE), avgM = mean(Body_mass_male_mean,
        na.rm = TRUE), avgBMD = mean(Mass_Dimorphism, na.rm = TRUE))
s
```


Acccording to Kamilar & Cooper’s (2013) dataset, what is the average male and female size, and body mass dimorphism of my two main study species (vervet monkeys, Chlorocebus pygerythrus; and woolly monkeys, Lagothrix lagotricha)? Which has a larger average female body mass? Which is more sexually dimorphic?

```{r}
cp <- s %>%
  filter(Binomial =="Chlorocebus pygerythrus")
ll <- s %>%
  filter(Binomial == "Lagothriz lagotricha")
cp
ll

```

Compare the body size of my two main study taxa at the Family level (i.e., Cercopithecidae vs. Atelidae) by plotting (using {ggplot2}) the body mass of males and females and sexual dimorphism. If you can, make the Cercopithecid boxes green, and the Atelid boxes purple.