---
title: "Module-6"
format: html
editor: visual
---

### Module 6

```{r} 
library(tidyverse)
```

### Challenge 1

# Load the Country-Data-2016 into a data frame varaible, d, and summarize the variables in that data frame. You can load the .csv file with a link or from local file so I'm going to try both.

# Load with link/curl
```{r}
library(curl)
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/Country-Data-2016.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
```

## OR
# Load from local file
```{r}
f <- "~/Desktop/Country-Data-2016.csv"
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
```

```{r}
summary(d)
```

```{r}
names(d)
```

# Create a new pop_density variable in your data frame which is population / area. What are the 10 most dense countries? The 10 least dense?

# Most
```{r}
d$pop_density <- d$population/d$area
d <- d[order(-d$pop_density), ]
d[1:10, ]
```

# Least
```{r}
d <- d[order(d$pop_density), ]
d[1:10, ]
```

# Extract data from the 20 largest countries into a new variable (I believe a new varaible refers to a new dataframe (NOT d)). What are the median area and population size of these countries?
```{r}
big_20 <- d[order(-d$area), ] [1:20,]
summary(big_20)
```

## OR
```{r}
mean(new$population, na.rm = TRUE)
mean(new$area, na.rm = TRUE)
```

# Extract data from all countries beginning with the letters “A” through “F”. What are the mean area and population size of these countries?
# grep looks for A-F anywhere in the name of the country
```{r}
a_to_f <- d[grep("[A-F]", d$country), ]
summary(a_to_f)
```

### Challenge 2

```{r}
par(mfrow = c(2, 3))
boxplot(d$population)
boxplot(log(d$population))
boxplot(d$area)
boxplot(log(d$area))
barplot(d$population)
barplot(d$area)
```

### Challenge 3




